US General Society Survey Analysis

### 2024-03-17 paper/paper.qmd
finsihed analysis and discussion for model 3

The United States of America is a democratic government which means that they hold election every 4 years[@usc_presidential_election]. The 2 parties that historically run against each other with the most votes are the Republican party and the Democratic Party. The republicans are often associated with conservative beliefs and values such as views opposing abortion and privatization to save their economy. The Democrats, on the other hand, are often associated with liberal views such as social welfare programs and higher taxes to support the government aids provided to citizens[@britannica_dem_vs_rep]. There are many other parties such as Libertarian Party, Green Party, Constitution Party and other independent candidates but because majority of the votes goes to the 2 parties the others are often over looked. The legal voting age is 18 in the states and you must be a registered voter in order to take part of the election which means non-citizens are not taken into account. Students are not as well. Also there seems to be a 66% voter turnout which means the remaining 34% decided not to vote [@DeSilver_2022]. The 2024 elections is also coming up this year meaning that analyzing this results may help us predict what the outcome of the election might be.

@@

### Aging Voters (2016)

In the following graphs we will be looking at the percentage of voters from each birth year that voted for Hillary Clinton, Donald Trump,and other candidates in the 2016 US Presidential election, we will also take a look at those who did not vote.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-hillary2016
#| fig-cap: "Comparing Percentage of Votes For Hillary Clinton and Year of Birth (2016)"

ces2020_votes <- ces2020 %>%
  filter(presvote16post %in% c("Hillary Clinton", "Donald Trump", "Other", "Did not vote"))

@@

#R value
hillary_data <- percentage_data_long %>% filter(Candidate == "HillaryClinton")

R_hillary <- cor(hillary_data$birthyr, hillary_data$Percentage)

plot_hillary <- ggplot(percentage_data_long %>% filter(Candidate == "HillaryClinton"), aes(x = birthyr, y = Percentage)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Percentage for Hillary Clinton", y = "Percentage (%)", x = "Birth Year") +
  theme_minimal()
plot_hillary

@@

In @fig-hillary2016 we can see that the younger the voters, the smaller percentage of their birth year seem to have voted for Hillary Clinton in the 2016 presidential election. This was rather surprising for us, as we expected young voters more likely to vote for Hillary Clinton and the Democratic Party, as younger people are often more progressive than their older counterparts. @fig-hillary shows us that there seems to be a significant correlation between age, birth year, and what percent of that birth year voted for Hillary with a correlation coefficient of $R = -0.94$, which is a rather strong negative relationship between our two variables.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-trump2016
#| fig-cap: "Comparing Percentage of Votes For Donald Trump and Year of Birth (2016)"

trump_data <- percentage_data_long %>% filter(Candidate == "DonaldTrump")

R_trump <- cor(trump_data$birthyr, trump_data$Percentage)

#Trum Plot
plot_trump <- ggplot(percentage_data_long %>% filter(Candidate == "DonaldTrump"), aes(x = birthyr, y = Percentage)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Percentage for Donald Trump", y = "Percentage (%)", x = "Birth Year") +
  theme_minimal()
plot_trump
```

While we look at the graph for Donald Trump in the 2016 election at first glance it does seem that the younger you are the more likey you were to have voted for trump in the 2016 election, but if we dig a bit deeper we can see that there seem to be several outliers on this graph, and the data points seem to somewhat be bouncing around a bit. This is confirmed when we look at our correlation coefficient $ R = 0.12$, so we can come to a conclsion that voting for Donald Trump in 2016 has a relatively weak positive correlation.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-other2016
#| fig-cap: "Comparing Percentage of Votes For Other Caniddates and Year of Birth (2016)"

other_data <- percentage_data_long %>% filter(Candidate == "Other")

R_other <- cor(other_data$birthyr, other_data$Percentage)

@@

@fig-other2016 does not seem to give us to much information about the age of those who voted for other candidates in the 2016 presidential elections, such as Gary Johnson and Jill Stein. While all these candidates received a combined less than 5% of the total vote, and less than 5% of the share of the vote from each birth year, we believe that transparency is incredibly important when doing data heavy research and while this may not seem as important as the graphs for Hillary Clinton and Donald Trump, this graph absolutely has a place in this paper.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-dnv2016
#| fig-cap: "Comparing Percentage of Non-Voters and Year of Birth (2016)"

dnv_data <- percentage_data_long %>% filter(Candidate == "Did not vote")

R_dnv <- cor(dnv_data$birthyr, dnv_data$Percentage)

@@

Here we will look at the members of the population that did not vote. @fig-dnv2016

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-biden2020
#| fig-cap: "Comparing Percentage of Votes For Joe Biden and Year of Birth (2020)"

@@

#Rvalue Biden
biden_data <- percentage_data_long %>% filter(Candidate == "Biden")

R_biden <- cor(biden_data$birthyr, biden_data$Percentage)

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-trump2020
#| fig-cap: "Comparing Percentage of Votes For Donald Trump Year of Birth (2020)"

@@

plot_trump_2

#rvalue Trump 2020

trump_2_data <- percentage_data_long %>% filter(Candidate == "Trump")

R_trump_2 <- cor(trump_2_data$birthyr, trump_2_data$Percentage)
```

@@

$$
\begin{aligned} 
y_i|\pi_i  &\sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \beta_0 + \beta_1 \times \mbox{gender}_i + \beta_2\times \mbox{immigration status}_i\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5)
\end{aligned}
$$

@@

political_preferences <-
  stan_glm(
    voted_for ~ gender + immstat,
    data = ces2020_reduced,
    family = binomial(link = "logit"),
    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
    prior_intercept = 
      normal(location = 0, scale = 2.5, autoscale = TRUE),
    seed = 302
  )

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of presidential election in 2020 based on gender and immigration status"
#| warning: false
modelsummary(
  list(
    "Support Biden" = political_preferences
  ),
  statistic = "mad"
  )
```

@@

## Third discussion point

@@

  It is near impossible to have data sets that are completely error-free and the same is true for this data-set. The data-set we used as analysis does not take into account the opinions of non-citizens and who they would have voted for. A lot of marginalized opinions are being left out when we clean the data and we only look for people that could vote. For example refugees may have a different opinion on who got elected in the 2020 presidential candidate as they come from a different aspect of life and have different priorities that need to be addressed such as housing and or health care. 
  Another weakness in our data-set is what occurs with a lot of data-sets, which is missing values. This data set also contained them and those needed to be cleaned so that our analysis would not sway. Speaking of missing values, insignificant values were ignored in some of the models as well. For, example some of the presidential candidates got so little votes that including them would have actually hindered our ability to see the trends so we had to group them in others. This was beneficial for this paper but  if we do indeed continue to analyse this paper for other purposes, this might cause problems.

  In terms of next steps for this paper, we look forward to predict the 2024 election results by comparing the what the general priorities of the population is. We could look at population by immigration status, gender, age, or their previous 2020 voting results.  If we are able to correctly predict the results based on the model continuously, it would help understand the priorities of american citizens; knowledge that would in turn help us help them.

@@

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

>>>>

The United States of America is a democratic government which means that they hold election every 4 years[@usc_presidential_election]. The 2 parties that historically run against each other with the most votes are the Republican party and the Democratic Party. The republicans are often associated with conservative beliefs and values such as views opposing abortion and privatization to save their economy. The Democrats, on the other hand, are often associated with liberal views such as social welfare programs and higher taxes to support the government aids provided to citizens[@britannica_dem_vs_rep]. There are many other parties such as Libertarian Party, Green Party, Constitution Party and other independent candidates but because majority of the votes goes to the 2 parties the others are often over looked. The legal voting age is 18 in the states and you must be a registered voter in order to take part of the election which means non-citizens are not taken into account. Students are not as well. Also there seems to be a 66% voter turnout which means the remaining 34% decided not to vote [@DeSilver_2022]. The 2024 elections is also coming up this year meaning that analyzing this results may help us predict what the outcome of the election might be. As for the results of the election in 2020, Donald Trump won the precidential title.

@@

### Voters in 2016

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-hillary2016
#| fig-cap: "Linear Regression comparing Percentage of Votes For Hillary Clinton and Year of Birth (2016)"

# only choose values "Hillary Clinton" and "Donald Trump"
ces2020_votes <- ces2020 %>%
  filter(presvote16post %in% c("Hillary Clinton", "Donald Trump", "Other", "Did not vote"))

@@

plot_hillary <- ggplot(percentage_data_long %>% filter(Candidate == "HillaryClinton"), aes(x = birthyr, y = Percentage)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Percentage for Hillary Clinton", y = "Percentage (%)") +
  theme_minimal()
plot_hillary

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-trump2016
#| fig-cap: "Linear Regression comparing Percentage of Votes For Donald Trump and Year of Birth (2016)"

#Trum Plot
plot_trump <- ggplot(percentage_data_long %>% filter(Candidate == "DonaldTrump"), aes(x = birthyr, y = Percentage)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Percentage for Donald Trump", y = "Percentage (%)") +
  theme_minimal()
plot_trump
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-other2016
#| fig-cap: "Linear Regression comparing Percentage of Votes For Other Caniddates and Year of Birth (2016)"

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-dnv2016
#| fig-cap: "Linear Regression comparing Percentage of Non-Voters and Year of Birth (2016)"

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-biden2020
#| fig-cap: "Linear Regression comparing Percentage of Votes For Joe Biden and Year of Birth (2020)"

@@

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-trump2020
#| fig-cap: "Linear Regression comparing Percentage of Votes For Donald Trump Year of Birth (2020)"

@@

plot_trump_2
```

@@

$$
\begin{aligned} 
y_i|\pi_i  &\sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) &= \beta_0 + \beta_1 \times \mbox{gender}_i + \beta_2\times \mbox{immigration status}_i + \beta_3\times \mbox{previous 2016 vote}_i\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5)
\end{aligned}
$$

@@

political_preferences <-
  stan_glm(
    voted_for ~ gender + immstat + presvote16post ,
    data = ces2020_reduced,
    family = binomial(link = "logit"),
    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
    prior_intercept = 
      normal(location = 0, scale = 2.5, autoscale = TRUE),
    seed = 302
  )

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults3
#| tbl-cap: "Explanatory models of presidential election in 2020 based on gender and immigration status"
#| warning: false
modelsummary(
  list(
    "Support Biden" = political_preferences
  ),
  statistic = "mad"
  )
```

The numbers on the first half of the table @tbl-modelresults3, right next to each variable name, shows us the coefficient related to each of the predictor variables. It tells us whether the respondent is expected to vote for Biden or not based on their immigration status, gender and who they voted for in the 2016 elections. The bracketed values are the errors associated with the coefficients so the smaller the number the less likely the model inaccurately came up with the coefficients. Notice that one of the variables is named intercept which essentially represents gender Female. We see that in "parent and I born in US but grandparent(s) immigrant", "gender male" and "gender female", the coefficients are negative meaning that they are not inclined to vote for Biden in the 2020 election. All other values are positive and we see for people who voted for Trump in the 2016 election, they have a higher coefficient with a relatively small inaccuracy rate.

@@

## Modeling immigration status, gender and 2016 voting results

   @tbl-modelresults3 was a very interesting plot to examine as they let us see the correlation between the election held in 2020 and other factors such as gender, immigration status, and the election results in 2016. Starting off we look at the negative values. We see that both males and females are less likely to vote for Biden as their coefficient is negative. The female has more of a negative value meaning that their priorities do not align with the view of the democratic party. This makes sense as we look at the news in the past years when we see the Biden government push for anti-abortion laws, the very law that a lot of women are opposed to as they argue they deserve to make that choice. As for the negative value for the males, there does not seem to be a direct answer to that. From here we see that it makes more sense to look at the large coefficients as the smaller values may be a result of many different things. A person is just not male but they may also be from a different ethnicity or religion whose priorities are different from the Biden government. America is a very diverse country so it makes sense that many people have different reasons to support and not support the republicans. The next largest value we notice is for people who were born in the US but have immigrant parents and people who are immigrants but not citizens. This makes sense as the Biden government actually reversed some of the immigration plans implemented by the previous government that harmed immigrants. Some of these laws were the travel ban put on Muslim-majority countries that resulted in many immigrants being unable to come to America to find a better future. It makes sense that children of people who immigrated and people who are immigrants themselves would prioritize such issues. Lastly, we look at the another coefficient found in the table which is the people who voted for Trump in the 2016 election. We see that it is very likely that people who voted for Trump in the last election, voted for Biden in this one. This may be due to the fact that there were many protests of the Trump administration and how they ran the country during their term. Some of the backlash that happened while Trump was president was the travel ban on Muslim-majority countries, reduced government regulation across various industries, and tax cuts from private corporations. These decisions may have led many people to not re-elect Trump for another 4 years as their president. The same is seen for people who voted for Evan McMullin (independent candidate) and other smaller parties. The reasons are hard to infer but it may be due to the fact that their candidate did not run again or their priorities and views changed from the party/person they voted for.

@@

  It is nearly impossible to have data sets that are completely error-free and the same is true for this data set. The data set we used as analysis does not take into account the opinions of non-citizens and who they would have voted for. A lot of marginalized opinions are being left out when we clean the data and we only look for people that could vote. For example, refugees may have a different opinion on who got elected in the 2020 presidential candidate as they come from a different aspect of life and have different priorities that need to be addressed such as housing and or health care. 
  Another weakness in our data set is what occurs with a lot of data sets, which is missing values. This data set also contained them and those needed to be cleaned so that our analysis would not sway. Speaking of missing values, insignificant values were ignored in some of the models as well. For, example some of the presidential candidates got so few votes that including them would have actually hindered our ability to see the trends so we had to group them into others. This was beneficial for this paper but if we do indeed continue to analyze this paper for other purposes, this might cause problems.

  In terms of the next steps for this paper, we look forward to predicting the 2024 election results by comparing the general priorities of the population. We could look at the population by immigration status, gender, age, or their previous 2020 voting results.  If we are able to correctly predict the results based on the model continuously, it would help us understand the priorities of American citizens; knowledge that would in turn help us help them.

@@

```{r}
#| echo: false
#| eval: true
#| label: fig-modelCred
#| fig-cap: "Credible intervals for predictors of support for Biden"
#| warning: false

modelplot(political_preferences, conf_level = 0.9) +
  labs(x = "90 per cent credibility interval")
```

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check.

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior.
